---
title: "Workshop1"
author: "maz"
format: html
editor: visual
---

## Workshop 1

```{r}
install.packages("ggplot2")
library(ggplot2)
```

```{r}
data(airquality)
```

```{r}
View(airquality)  # Opens the dataset in a spreadsheet-like viewer
head(airquality)  # Displays the first few rows of the dataset

```

```{r}
summary(airquality)  # Gives a statistical summary of the dataset
str(airquality)      # Shows the structure of the dataset

```

1

```{r}
mean(airquality$Ozone, na.rm = TRUE)
mean(airquality$Solar.R, na.rm = TRUE)
```

**Explanation**: This command calculates the mean (average) values for the 'Ozone' and 'Solar.R' columns, excluding any missing values (**`NA`**) in the dataset.

2

```{r}
mean_Ozone <- mean(airquality$Ozone, na.rm = TRUE)

mean_SolarR <- mean(airquality$Solar.R, na.rm = TRUE)

print(paste("Mean Ozone Level:", mean_Ozone))
print(paste("Mean Solar Radiation:", mean_SolarR))
```

-   **R Command**: Same as above, with **`na.rm = TRUE`**.

-   **Explanation**: The **`na.rm = TRUE`** argument tells R to ignore the **`NA`** (Not Available) values while calculating the mean.

3

```{r}
var(airquality$Ozone, na.rm = TRUE)
sd(airquality$Ozone, na.rm = TRUE)

```

**Explanation**: These commands calculate the variance and standard deviation for the 'Ozone' variable, providing insights into its dispersion or spread.

4

```{r}
median(airquality$Ozone, na.rm = TRUE)

```

**Explanation**: This computes the median (middle value) for the 'Ozone' variable, giving a sense of the central tendency.

5

```{r}
summary(airquality$Ozone)

```

**Explanation**: The **`summary`** function in R provides the five number summary (minimum, first quartile, median, third quartile, maximum) for the 'Ozone' variable.

6

```{r}
hist(airquality$Ozone, main="Ozone Histogram", xlab="Ozone")
boxplot(airquality$Ozone, main="Ozone Box Plot")

```

**Explanation**: This command converts the 'Month' variable into a factor, which is R's way of handling categorical data.

7

```{r}
hist(airquality$Ozone, main="Ozone Histogram", xlab="Ozone")
boxplot(airquality$Ozone, main="Ozone Box Plot")
```

**Explanation**: These commands create a histogram and a box plot for the 'Ozone' variable, helping visualize the distribution and spread of the data.

## Extra

```{r}
# Install necessary libraries
install.packages("shiny")
install.packages("plotly")  # For interactive plots
install.packages("dplyr")   # For data manipulation
install.packages("DT")      # For interactive tables

# Load the libraries
library(shiny)
library(plotly)
library(dplyr)
library(DT)


```

```{r}
data(airquality)
```

#### UI (User Interface)

The UI defines the layout and appearance of your app:

```{r}
ui <- fluidPage(
    titlePanel("Air Quality Data Visualization"),
    
    sidebarLayout(
        sidebarPanel(
            selectInput("xvar", "Select X Variable:", choices = names(airquality)),
            selectInput("yvar", "Select Y Variable:", choices = names(airquality)),
            selectInput("numvar", "Select Numeric Variable for Histogram/Density Plot:", 
                        choices = c("Ozone", "Solar.R", "Wind", "Temp")),
            selectInput("chartType", "Select Chart Type:",
                        choices = c("Bar Chart", "Histogram", "Box Plot", "Scatter Plot", 
                                    "Line Chart", "Time Series", "Area Chart", "Density Plot"))
        ),
        
        mainPanel(
            plotOutput("plot")
        )
    )
)



```

#### Server Function

The server function contains the app's logic:

```{r}
server <- function(input, output) {
    
    output$plot <- renderPlot({
        data <- airquality

        # Base plot, using a conditional for histogram and density plot
        plot <- if (input$chartType %in% c("Histogram", "Density Plot")) {
            ggplot(data, aes_string(x = input$numvar))
        } else {
            ggplot(data, aes_string(x = input$xvar, y = input$yvar))
        }

        # Add appropriate geom based on selected chart type
        switch(input$chartType,
               "Bar Chart" = plot + geom_bar(stat = "identity"),
               "Histogram" = plot + geom_histogram(binwidth = 1, na.rm = TRUE),
               "Box Plot" = plot + geom_boxplot(),
               "Scatter Plot" = plot + geom_point(),
               "Line Chart" = plot + geom_line(),
               "Time Series" = plot + geom_line(),
               "Area Chart" = plot + geom_area(),
               "Density Plot" = plot + geom_density()
        )
    })
}



```

#### Run the App

To start the app, use:

```{r}
shinyApp(ui = ui, server = server)

```
