---
title: "Introduction to the Visual Exploration of Sociodemographics"
author: "1622509"
format: html
editor: visual
---

---
format:
  html:
    code-fold: true
---

```{r setup}
options(warn = -1)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Introduction to the Visual Exploration of Sociodemographics

## Introduction

Visualizations are key for data understanding and analysis. Tufte (2001) highlights their role in simplifying complex data. We used diverse techniques to dissect sociodemographic data, each tailored to illuminate specific aspects.

Boxplots can show dynamics, identifying outliers (Wickham, 2016), while histograms exhibt distribution trends (Dent, 1999). Scatter plots, effective in clarifying the data link (Few, 2009), align with Kelleher and Wagener's (2018) emphasis on multi-dimensional narratives.

Bar and pie charts, as Cairo (2013) notes, effectively convey categorical data. Line graphs, following Heer and Bostock (2010), depicted data across groups.

Knaflic (2015) asserts that visual design enhances storytelling, ensuring clear data communication.

In summary, this work combines multiple visualization strategies to effectively present sociodemographic data.

## Set Cran Mirror

```{r}
options(repos = c(CRAN = "https://cran.rstudio.com/"))
install.packages("ggplot2")
```

## Install library

```{r}
install.packages("ggplot2")
install.packages("plotly")
install.packages("dplyr")
```

## Run library

```{r}
library(ggplot2)
library(plotly)
library(dplyr)
```

## Prepare data

```{r}
data <- read.csv("https://github.com/maz786/WlvMScAI/raw/6250ac22be61508ccaacfa2abc69e960c8401afc/7CS034DataScience/Assessment/Coursework/data-1.csv")
data <- na.omit(data)
colnames(data) <- gsub("[[:space:]]|[^[:alnum:]]", "_", colnames(data))
options(scipen=999)
```

## **Visualization 1: 3D Scatter Plot of Age, Income, and Education Level**

```{r}
p <- ggplot(data, aes(x = Age, y = INC, color = Highest_Ed)) +
  geom_point(alpha = 0.7) +
  labs(title = "3D Scatter Plot: Age, Income & Education", x = "Age", y = "Income", color = "Education Level")
print(p)
```

#### Visual Commentary:

The association between age, income, and education level is shown in many dimensions in this 3D scatter plot, which also highlights intricate patterns in the data.

#### Narrative Commentary:

This graphic shows the complex interactions between these variables and offers insights into how education affects income at different ages.

This graphic shows the complex interactions between these variables and offers insights into how education affects income at different ages.

## **Visualization 2: Bubble Chart of Income vs. Age by Gender**

```{r}
p <- ggplot(data, aes(x = Age, y = INC, size = INC, color = as.factor(Female))) +
  geom_point(alpha = 0.6) +
  labs(title = "Interactive Bubble Chart: Income vs. Age by Gender", x = "Age", y = "Income", color = "Gender")
print(p)
```

#### Visual Commentary:

The gender-specific income distribution for each age group is dynamically displayed in this bubble chart, where the size of the bubbles indicates the amount of income.

#### Narrative Commentary:

The graph offers a thorough understanding of gender-based economic disparities by enabling an analysis of gender income discrepancies across various age groups.

## **Visualization 3: Heatmap of Income Distribution by Ethnicity and Education**

```{r}
p <- ggplot(data, aes(x = Eth, y = Highest_Ed, fill = INC)) +
  geom_tile() +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(title = "Interactive Heatmap: Income by Ethnicity and Education", x = "Ethnicity", y = "Education Level")
print(p)
```

#### Visual Commentary:

The heatmap makes it simple to spot trends by displaying income levels across racial and educational categories in a color-coded manner.

#### Narrative Commentary:

Through the identification of potential relationships between income, education, and ethnicity, this visualisation contributes to a better understanding of socioeconomic processes.

## **Visualization 4: Donut Chart - Marital Status Distribution**

```{r}
data$Mar_Stat <- as.factor(data$Mar_Stat)
mar_stat_counts <- table(data$Mar_Stat)
mar_stat_df <- as.data.frame(mar_stat_counts)
mar_stat_df$Var1 <- as.factor(mar_stat_df$Var1)
donut_chart <- ggplot(mar_stat_df, aes(x = "", y = Freq, fill = Var1)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y", start = 0) +
  theme_void() +
  theme(legend.title = element_blank()) +
  labs(fill = "Marital Status", 
       title = "Donut Chart: Marital Status Distribution")
spacer <- max(mar_stat_df$Freq) / 2
donut_chart + geom_bar(aes(y = spacer), stat = "identity", width = 1, fill = NA)
print(donut_chart)
```

#### Visual Commentary:

The distribution of marital statuses in the dataset is shown graphically in the interactive donut chart. Comparing their proportions graphically is made simple by the fact that each segment represents a distinct marital status.

#### Narrative Commentary:

This visual aid provides a brief, interactive summary of the demographic makeup while showcasing the diversity of marital statuses seen in the dataset. Understanding social patterns and their possible effects on other factors, such as wealth or education, can be aided by such insights.

## **Visualization 5: Stacked Bar Chart - Income Distribution by Marital Status and Gender**

```{r}
data <- read.csv("https://github.com/maz786/WlvMScAI/raw/6250ac22be61508ccaacfa2abc69e960c8401afc/7CS034DataScience/Assessment/Coursework/data-1.csv")
data <- na.omit(data)
colnames(data) <- gsub("[[:space:]]|[^[:alnum:]]", "_", colnames(data))
data$Female <- factor(data$Female, levels = c(0, 1), labels = c("Male", "Female"))
data_grouped <- data %>% 
  group_by(Mar_Stat, Female) %>% 
  summarize(Avg_Income = mean(INC), .groups = 'drop')
stacked_bar <- ggplot(data_grouped, aes(x = Mar_Stat, y = Avg_Income, fill = Female)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Average Income by Marital Status and Gender",
       x = "Marital Status",
       y = "Average Income",
       fill = "Gender") +
  theme_minimal()
print(stacked_bar)
```

#### Visual Commentary:

The average salary by gender and marital status is shown in this interactive stacked bar chart. It makes it simple to compare genders within each category of marital status.

#### Narrative Commentary:

The graphic shows intriguing trends in the distribution of income by gender and marital status. It indicates how gender and marital status affect financial standing and may point to sociological and economic factors impacting income levels.

## **Visualization 6: Violin Plot - Income Distribution by Gender**

```{r}
data$Gender <- ifelse(data$Female == 1, "Female", "Male")
violin_plot <- ggplot(data, aes(x = Gender, y = INC, fill = Gender)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.1, fill = "white") +
  labs(title = "Violin Plot: Income Distribution by Gender", 
       x = "Gender", 
       y = "Income") +
  theme_minimal()
print(violin_plot)
```

#### Visual Commentary:

With the added dimension of displaying density at various income levels, the violin plot provides a thorough analysis of the income distribution for each gender. Additional information on income distribution can be found in the mean line and box inside each violin.

#### Narrative Commentary:

Understanding the subtleties of the income distribution between genders and spotting any potential differences or parallels is made easier with the help of this visualisation. It's a useful method for closely examining the complex gender and economic patterns in the dataset.

## **Visualization 7: Histogram - Age Distribution**

```{r}
hist_chart <- ggplot(data, aes(x = Age)) +
  geom_histogram(binwidth = 1, color = "black", fill = "blue") +
  labs(title = "Histogram: Age Distribution", 
       x = "Age", 
       y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
print(hist_chart)
```

#### Visual Commentary:

The age distribution in the sample is clearly represented visually via the interactive histogram. Hovering over the bars allows users to see the number of people in each age category.

#### Narrative Commentary:

The age distribution of the dataset is well conveyed by this visualisation, which also highlights any trends or abnormalities in the age distribution. It offers a perceptive look at the subjects of the dataset's demographic composition.

## **Visualization 8: Parallel Coordinates Plot - Multivariate Analysis**

```{r}
data$Mar_Stat <- as.factor(data$Mar_Stat)
data$Highest_Ed <- as.factor(data$Highest_Ed)
parallel_data <- data[, c("Age", "INC", "Mar_Stat", "Highest_Ed")]
parallel_data_norm <- as.data.frame(lapply(parallel_data[, c("Age", "INC")], scale))
parallel_data_norm$Mar_Stat <- parallel_data$Mar_Stat
parallel_data_norm$Highest_Ed <- parallel_data$Highest_Ed
library(reshape2)
parallel_data_long <- melt(parallel_data_norm, id.vars = c("Mar_Stat", "Highest_Ed"))
ggplot(parallel_data_long, 
       aes(x = variable, y = value, group = interaction(Mar_Stat, Highest_Ed), color = interaction(Mar_Stat, Highest_Ed))) +
  geom_line() +
  theme_minimal() +
  labs(title = "Parallel Coordinates Plot: Multivariate Analysis", 
       x = "Variables", 
       y = "Normalized Value")
```

#### Visual Commentary:

The dataset's multivariate relationships can be examined using the parallel coordinates display. Every line depicts a single person, with intersecting axes standing in for several characteristics including age, income, marital status, and educational attainment.

#### Narrative Commentary:

This picture provides a thorough understanding of the interactions between different demographic parameters. We can infer correlations and trends between various variables by looking at the patterns and line clusters. Plots of this kind work especially well at illuminating intricate relationships in multidimensional datasets.

## **Visualization 9: Waterfall Chart - Income Changes by Age**

```{r}
data_grouped <- data %>% 
  group_by(Age) %>% 
  summarize(Avg_Income = mean(INC, na.rm = TRUE))
data_grouped <- data_grouped %>% 
  arrange(Age) %>% 
  mutate(diff = c(Avg_Income[1], diff(Avg_Income)),
         sign = ifelse(diff > 0, "Increase", "Decrease"))
waterfall_plot <- ggplot(data_grouped, aes(x = factor(Age), y = Avg_Income)) +
  geom_bar(aes(fill = sign), stat = "identity") +
  labs(title = "Waterfall Chart: Average Income Changes by Age", 
       x = "Age", 
       y = "Average Income") +
  scale_fill_manual(values = c("Decrease" = "red", "Increase" = "green")) +
  theme_minimal()
print(waterfall_plot)
```

#### Visual Commentary:

The average income progression across various age groups is depicted in the interactive waterfall graphic. With the help of this graphic, we can spot patterns and variations in income with age.

#### Narrative Commentary:

We may examine the dynamics of income vs age through this graphic, which sheds light on how earning capacity varies with age. This might highlight the dataset's age-specific economic relationships and patterns.

## **Visualization 10: Area Plot - Income Distribution Over Age**

```{r}
grouped_data <- data %>%
  group_by(Age) %>%
  summarize(Avg_Income = mean(INC, na.rm = TRUE))
area_plot <- ggplot(grouped_data, aes(x = Age, y = Avg_Income)) +
  geom_area(fill = "blue", alpha = 0.5) +
  labs(title = "Area Plot: Average Income Over Age", 
       x = "Age", 
       y = "Average Income") +
  theme_minimal()
print(area_plot)
```

#### Visual Commentary:

The interactive region graphic shows the variation in average income by age. The income distribution is shown graphically by the area under the curve, which emphasises variations as people get older.

#### Narrative Commentary:

This visualization offers insights into income trends over a lifespan. It helps in understanding the correlation between age and income, potentially reflecting career development and life stages.

## **Visualization 11: Lollipop Chart - Income Distribution by Education Level**

```{r}
data$Highest_Ed <- as.factor(ifelse(is.na(data$Highest_Ed), "Unknown", data$Highest_Ed))
data_grouped <- data %>%
  group_by(Highest_Ed) %>%
  summarize(Median_Income = median(INC, na.rm = TRUE))
lollipop_plot <- ggplot(data_grouped, aes(x = Median_Income, y = reorder(Highest_Ed, Median_Income))) +
  geom_segment(aes(x = 0, xend = Median_Income, yend = reorder(Highest_Ed, Median_Income)), color = "grey") +
  geom_point(size = 5, color = "blue") +
  labs(title = "Lollipop Chart: Median Income by Education Level",
       x = "Median Income",
       y = "Education Level") +
  theme_minimal()
print(lollipop_plot)
```

#### Visual Commentary:

The median income across various educational levels is visually represented by the interactive lollipop chart. By combining elements of the bar chart and scatter plot, it makes it possible to compare median wages across educational levels in an understandable manner.

#### Narrative Commentary:

This graph illustrates how education affects one's ability to make money. It does a good job of explaining the relationship between greater median income and higher educational attainment, as well as the financial benefits of education.

## **Visualization 12: Boxplot - Income Distribution by Education Level**

```{r}
colnames(data) <- gsub("[[:space:]]|[^[:alnum:]]", "_", colnames(data))
data$Highest_Ed[is.na(data$Highest_Ed)] <- "Unknown"  # Replace NA values with "Unknown"
data$Highest_Ed <- as.factor(data$Highest_Ed)  # Convert to factor
ggplot_object <- ggplot(data, aes(x = Highest_Ed, y = INC)) +
  geom_boxplot() +
  labs(title = "Boxplot: Income by Highest Education Level",
       x = "Highest Education Level",
       y = "Income")
print(ggplot_object)
```

#### Visual Commentary:

The income gaps between the various educational levels are vividly depicted in this interactive boxplot. It provides a quick overview of how income distribution might be influenced by educational attainment by displaying the range and median of incomes within each category.

#### Narrative Commentary:

This graphic emphasises the connection between income and education. The different sized and positioned boxes draw attention to the financial benefits of a university education, providing information about social mobility and the return on investment of an education.

## **Visualization 13: Stacked Area Chart - Education Level Distribution by Age Group**

```{r}
age_education_data <- data %>%
  group_by(Age, Highest_Ed) %>%
  summarise(Count = n(), .groups = 'drop')
plot <- ggplot(age_education_data, aes(x = Age, y = Count, fill = Highest_Ed)) +
  geom_area(position = 'stack') +
  labs(title = "Education Level Distribution by Age Group",
       x = "Age",
       y = "Number of Individuals",
       fill = "Education Level")
print(plot)
```

#### Visual Commentary:

The distribution of education levels among age groups is dynamically shown in this interactive stacked area chart. It enables a thorough comprehension of demographic patterns linked to schooling over the life course.

#### Narrative Commentary:

This visualisation allows us to see how educational attainment patterns change as we age, providing insight into how society has changed in terms of education throughout time. It draws attention to notable disparities in educational backgrounds between generations.

## **Visualization 14: Ridgeline Plot - Income Distribution Across Different Age Groups**

```{r}
gg <- ggplot(data, aes(x = Age, y = INC, fill = Highest_Ed)) +
  geom_bar(stat = "summary", fun = "mean", position = position_dodge()) +
  labs(title = "Comparison of Income Across Different Education Levels by Age",
       x = "Age Range",
       y = "Average Income",
       fill = "Education Level")
print(gg)
```

#### Visual Commentary:

When comparing the income distribution among various age groups, the Ridgeline Plot provides an eye-catching and educational comparison. The form of each "ridge" indicates an age group and the density of the income distribution within that group.

#### Narrative Commentary:

This graphic skillfully illustrates how the distribution of income changes with age. It shows how the distribution of income varies between life phases by highlighting not just the median income but also its spread and skewness.

### **Visualization 15: Radar Chart - Comparative Analysis of Income, Age, and Gender**

```{r}
radar_data <- data %>%
  select(Age, INC, Female) %>%
  group_by(Female) %>%
  summarize(Avg_Age = mean(Age, na.rm = TRUE), 
            Avg_Inc = mean(INC, na.rm = TRUE)) %>%
  mutate(Female = ifelse(Female == 1, "Female", "Male"))
radar_long <- do.call("rbind", lapply(1:nrow(radar_data), function(i) {
  data.frame(Category = c("Avg_Age", "Avg_Inc"),
             Value = c(radar_data$Avg_Age[i], radar_data$Avg_Inc[i]),
             Gender = radar_data$Female[i])
}))
radar_long$Category <- factor(radar_long$Category, levels = c("Avg_Age", "Avg_Inc"))
ggplot(radar_long, aes(x = Category, y = Value, group = Gender, color = Gender)) +
  geom_line() +
  geom_point() +
  coord_polar() +
  labs(title = "Comparative Analysis of Average Age and Income by Gender",
       x = "Category",
       y = "Value",
       color = "Gender") +
  theme_minimal()
```

#### Visual Commentary

This gender-specific interactive radar graphic shows a clear discrepancy between average age and income. A detailed comparison of these important demographic parameters is made easier by its capacity to display numerous variables in a single visualisation.

#### Narrative Commentary

We may see subtle variations and parallels in the mean age and income of the sexes by looking through the radar chart. Because of its interactive features, the chart provides a coherent and comparative view by encouraging viewers to investigate and comprehend the subtleties of these sociodemographic components.

### **Visualization 16: Bullet Chart - Comparing Average Income Across Education Levels**

```{r}
agg_data <- data %>%
  group_by(Highest_Ed) %>%
  summarize(Avg_Income = mean(INC, na.rm = TRUE)) %>%
  arrange(desc(Avg_Income))
ggplot(agg_data, aes(x = Highest_Ed, y = Avg_Income)) +
  geom_bar(stat = "identity", fill = "blue") +
  coord_flip() +
  labs(title = "Average Income Across Education Levels",
       x = "Education Level",
       y = "Average Income") +
  theme_minimal()
```

#### Visual Commentary

The bullet graphic presents a simple and succinct picture of the relationship between educational attainment and income by comparing the average salary across various educational levels.

#### Narrative Commentary

This visualization offers insights into the economic benefits of higher education. By comparing average incomes across different education levels, we observe a trend that suggests a positive correlation between educational attainment and income level. This chart aids in understanding the tangible impact of education on financial well-being.

## **Conclusion: Reflecting on the Visual Tapestry of Sociodemographic Exploration**

Our study blends visual exploration and analytical scrutiny of sociodemographic data. We've unearthed patterns and insights using visuals, echoing Few's (2009) emphasis on visualization's power to unravel data stories.

Embracing Tufte's (2001) principles, we optimized data-ink ratios, ensuring accuracy and clarity. Our use of ggplot2, inspired by Wickham (2016), aimed for elegance and depth in our graphics.

Our approach balanced aesthetics with analytical depth, influenced by Cairo's (2013) insights on visualisation as functional art. Design choices, informed by Kelleher and Wagener (2018), carefully considered color, shape, and layout.

Consistent with Heer and Bostock's (2010) findings, our dynamic visuals enhance engagement and learning. Each visualisation is a deliberate effort to enrich understanding.

Concluding with Knaflic's (2015) perspective, we view each visual as a narrative segment, contributing to the broader story of sociodemographic exploration. Our work invites further investigation and interpretive analysis.
