---
title: "Introduction to the Visual Exploration of Sociodemographics"
author: "1622509"
format: 
  html:
    self_contained: true
    code-fold: true
editor: visual
---

```{r setup}
#| include: false
options(warn = -1)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Setup (Set Cran Mirror, install libraries, run libararies and prepare data)

```{r}
options(repos = c(CRAN = "https://cran.rstudio.com/"))
install.packages("ggplot2")
install.packages("plotly")
install.packages("dplyr")
library(ggplot2)
library(plotly)
library(dplyr)
data <- read.csv("https://github.com/maz786/WlvMScAI/raw/6250ac22be61508ccaacfa2abc69e960c8401afc/7CS034DataScience/Assessment/Coursework/data-1.csv")
data <- na.omit(data)
colnames(data) <- gsub("[[:space:]]|[^[:alnum:]]", "_", colnames(data))
options(scipen=999)
```

## Introduction to the Visual Exploration of Sociodemographics

In our study, we adopt a range of visualization techniques to clarify complex sociodemographic data, resonating with Tufte's (2001) principles of simplifying data through visualization. We utilize Interactive 3D Scatter Plots, Bubble Charts, Heatmaps, Donut Charts, Stacked Bar Charts, Histograms, Parallel Coordinates Plots, Area Plots, Boxplots, and Stacked Area Charts. These methods align with recommendations from experts like Wickham (2016) for boxplots to identify outliers, Dent (1999) for histograms to show distribution trends, Few (2009) for scatter plots to clarify data links, and Kelleher and Wagener (2018) for a multidimensional approach. Cairo (2013) underscores the effectiveness of bar and pie charts in presenting categorical data, while Heer and Bostock (2010) advocate for line graphs to compare data across groups. Our approach, in line with Knaflic (2015)'s emphasis on visual design, integrates these diverse visualizations to effectively present sociodemographic data.

## **Visualization 1: Interactive 3D Scatter Plot of Age, Income, and Education Level**

```{r}
p <- ggplot(data, aes(x = Age, y = INC, color = Highest_Ed)) +
  geom_point(alpha = 0.7) +
  labs(title = "3D Scatter Plot: Age, Income & Education", x = "Age", y = "Income", color = "Education Level")
ggplotly(p) %>% layout(scene = list(xaxis = list(title = 'Age'), yaxis = list(title = 'Income'), zaxis = list(title = 'Education Level')))
```

The 3D scatter plot reveals the multifaceted relationship between age, income, and education, highlighting complex patterns and providing insights into the influence of education on income across various ages

## **Visualization 2: Interactive Bubble Chart of Income vs. Age by Gender**

```{r}
p <- ggplot(data, aes(x = Age, y = INC, size = INC, color = as.factor(Female))) +
  geom_point(alpha = 0.6) +
  labs(title = "Interactive Bubble Chart: Income vs. Age by Gender", x = "Age", y = "Income", color = "Gender")
ggplotly(p)
```

This bubble chart dynamically illustrates gender-specific income distribution by age, with bubble size representing income levels, and elucidates gender-based economic disparities across different ages.

## **Visualization 3: Interactive Heatmap of Income Distribution by Ethnicity and Education**

```{r}
p <- ggplot(data, aes(x = Eth, y = Highest_Ed, fill = INC)) +
  geom_tile() +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(title = "Interactive Heatmap: Income by Ethnicity and Education", x = "Ethnicity", y = "Education Level")
ggplotly(p)
```

The heatmap color-codes income across racial and educational groups, simplifying the identification of trends and potential relationships among income, education, and ethnicity, aiding in the comprehension of socioeconomic dynamics.

## **Visualization 4: Interactive Donut Chart - Marital Status Distribution**

```{r}
mar_stat_counts <- table(data$Mar_Stat)
mar_stat_df <- as.data.frame(mar_stat_counts)
donut_chart <- plot_ly(data = mar_stat_df, labels = ~Var1, values = ~Freq, type = 'pie', hole = 0.5)
donut_chart <- donut_chart %>% layout(title = "Interactive Donut Chart: Marital Status Distribution", 
                                      showlegend = TRUE, 
                                      xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
                                      yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE))

# Display the plot
donut_chart
```

The interactive donut chart graphically represents the distribution of marital statuses, with each segment denoting a different status, offering a quick, interactive demographic overview and highlighting the variety within the dataset. It aids in understanding social patterns and their influence on aspects like wealth or education.

## **Visualization 5: Interactive Stacked Bar Chart - Income Distribution by Marital Status and Gender**

```{r}
data <- read.csv("https://github.com/maz786/WlvMScAI/raw/6250ac22be61508ccaacfa2abc69e960c8401afc/7CS034DataScience/Assessment/Coursework/data-1.csv")
data <- na.omit(data)
colnames(data) <- gsub("[[:space:]]|[^[:alnum:]]", "_", colnames(data))
data$Female <- factor(data$Female, levels = c(0, 1), labels = c("Male", "Female"))
data_grouped <- data %>% 
  group_by(Mar_Stat, Female) %>% 
  summarize(Avg_Income = mean(INC))
stacked_bar <- plot_ly(data_grouped, x = ~Mar_Stat, y = ~Avg_Income, type = 'bar', color = ~Female, colors = c('#FFD700', '#00BFFF'))
stacked_bar <- stacked_bar %>% layout(title = "Interactive Stacked Bar Chart: Average Income by Marital Status and Gender",
                                      barmode = 'stack',
                                      xaxis = list(title = "Marital Status"),
                                      yaxis = list(title = "Average Income"))
stacked_bar
```

The average salary by gender and marital status is shown in this interactive stacked bar chart. It makes it simple to compare genders within each category of marital status. The graphic shows intriguing trends in the distribution of income by gender and marital status. It indicates how gender and marital status affect financial standing and may point to sociological and economic factors impacting income levels.

## **Visualization 6: Interactive Histogram - Age Distribution**

```{r}
hist_chart <- plot_ly(data, x = ~Age, type = "histogram")
hist_chart <- hist_chart %>% layout(title = "Interactive Histogram: Age Distribution", 
                                    xaxis = list(title = "Age"),
                                    yaxis = list(title = "Count"),
                                    bargap = 0.05)
hist_chart
```

The interactive histogram vividly displays the sample's age distribution, with hover functionality revealing counts per age category. It effectively illustrates demographic composition, highlighting trends or anomalies in age distribution within the dataset.

## **Visualization 7: Interactive Parallel Coordinates Plot - Multivariate Analysis**

```{r}
data <- read.csv("https://github.com/maz786/WlvMScAI/raw/6250ac22be61508ccaacfa2abc69e960c8401afc/7CS034DataScience/Assessment/Coursework/data-1.csv")
data <- na.omit(data)
colnames(data) <- gsub("[[:space:]]|[^[:alnum:]]", "_", colnames(data))
data$Mar_Stat <- as.factor(data$Mar_Stat)
data$Highest_Ed <- as.factor(data$Highest_Ed)
parallel_data <- data[, c("Age", "INC", "Mar_Stat", "Highest_Ed")]
parallel_plot <- plot_ly(type = 'parcoords',
                         line = list(color = unlist(parallel_data$INC),
                                     colorscale = list(c(0, 'blue'), c(1, 'gold'))),
                         dimensions = list(
                           list(range = c(min(parallel_data$Age), max(parallel_data$Age)),
                                label = 'Age', values = parallel_data$Age),
                           list(range = c(min(parallel_data$INC), max(parallel_data$INC)),
                                label = 'Income', values = parallel_data$INC),
                           list(tickvals = 1:length(levels(parallel_data$Mar_Stat)),
                                ticktext = levels(parallel_data$Mar_Stat),
                                label = 'Marital Status', values = as.integer(parallel_data$Mar_Stat)),
                           list(tickvals = 1:length(levels(parallel_data$Highest_Ed)),
                                ticktext = levels(parallel_data$Highest_Ed),
                                label = 'Highest Education', values = as.integer(parallel_data$Highest_Ed))
                         ))
parallel_plot <- parallel_plot %>%
  layout(title = "Interactive Parallel Coordinates Plot: Multivariate Analysis")
parallel_plot
```

The parallel coordinates display allows for examination of the dataset's multivariate relationships, with each line representing an individual and intersecting axes for age, income, marital status, and education. It reveals interactions among demographic factors and helps infer correlations and trends, effectively illuminating complex relationships in multidimensional datasets.

## **Visualization 8: Interactive Area Plot - Income Distribution Over Age**

```{r}
grouped_data <- data %>%
  group_by(Age) %>%
  summarize(Avg_Income = mean(INC))
area_plot <- plot_ly(grouped_data, x = ~Age, y = ~Avg_Income, type = 'scatter', mode = 'lines', fill = 'tozeroy')
area_plot <- area_plot %>% layout(title = "Interactive Area Plot: Average Income Over Age")
area_plot
```

The interactive waterfall chart shows average income progression by age, highlighting patterns and variations in earnings over time. It allows for analysis of age-related income dynamics, emphasizing the dataset's age-specific economic trends.

## **Visualization 9: Interactive Boxplot - Income Distribution by Education Level**

```{r}
ggplot_object <- ggplot(data, aes(x = Highest_Ed, y = INC)) +
  geom_boxplot()
interactive_plot <- ggplotly(ggplot_object)
interactive_plot
```

The interactive boxplot vividly illustrates income disparities across educational levels, showing ranges and medians to reflect how education impacts income distribution. It highlights the financial rewards of higher education and informs on social mobility and educational ROI.

## **Visualization 10: Interactive Stacked Area Chart - Education Level Distribution by Age Group**

```{r}
age_education_data <- data %>%
  group_by(Age, Highest_Ed) %>%
  summarise(Count = n(), .groups = 'drop')
plot <- ggplot(age_education_data, aes(x = Age, y = Count, fill = Highest_Ed)) +
  geom_area(position = 'stack') +
  labs(title = "Education Level Distribution by Age Group",
       x = "Age",
       y = "Number of Individuals",
       fill = "Education Level")
fig <- ggplotly(plot)
fig
```

The interactive stacked area chart dynamically shows the distribution of education levels among age groups, offering insight into educational attainment over the lifespan and societal shifts in education over time, while highlighting generational educational disparities.

## **Conclusion: Reflecting on the Visual Tapestry of Sociodemographic Exploration**

Our study integrates visual and analytical methods to explore sociodemographic data, adhering to the principles of clarity and storytelling as outlined by Few (2009) and Tufte (2001). By utilizing ggplot2, we infused our visualizations with aesthetic depth in line with Wickham's (2016) recommendations, while also incorporating Cairo's (2013) concept of functional art to ensure our graphics are both attractive and informative. Thoughtful design choices, inspired by Kelleher and Wagener (2018), and interactive elements guided by Heer and Bostock's (2010) insights, enhanced user engagement. Finally, our visuals are framed within Knaflic's (2015) narrative approach, inviting deeper analysis and interpretation.
